Statistical dynamics of SARS-CoV-2 as a differential game

### 2020-07-23 paper/paper.tex
First draft.

\hypersetup
{
  pdftitle   = {Marginal dynamics of social distancing in COVID-19},
  pdfauthor  = {Chris von Csefalvay}
}

\title{Marginal dynamics of social distancing in COVID-19}
\author{Chris von Csefalvay}

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/fig1-odes}
	\caption{Some quantitative solutions for the SIR model's population dynamics over values of $R_0$ between 1.5 and 6.5, and values of $\gamma$ between $1/4.6$ and $1/12.4$ over a base population of 10,000 and a seed population of 1\% infected initially. For each plot, $\beta$ is inferred from an $R_0$ value of 2.67, based on Liu et al. (2020),\cite{liu2020reproductive} and the $\gamma$ parameter. The susceptible population is displayed in blue, while infected/infectious cases are marked in red and recovered cases in green.}
	\label{fig:ode_solutions}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/fig3-SIR-by-delta}
	\caption{Contour diagrams of numeric solutions for susceptible, infectious and removed ($S$, $I$ and $R$) compartment sizes for a base population of 10,000 individuals with a seed population of 1\% infected, under the assumption of an $R_0$ of 2.67 and $\gamma$ of $\frac{1}{8.5}$. As this figure indicates, the 'critical mass' of social distancing takes place in the $\delta$ range of 0 to 0.4, and thus even modest increases in social distancing participation at low levels can make a significant difference in the number of infectious cases.}
	\label{fig:fig3-SIR-by-delta}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/marginal_utility}
	\caption{Marginal utility of social distancing in a population $P$ of $\delta(P)$ adherence over time. The marginal utility is defined as the vertical component of the gradient of infected individuals. The plot draws on a base population of 10,000 individuals with a seed population of 1\% infected, under the assumption of an $R_0$ of 2.67 and $\gamma$ of $\frac{1}{8.5}$.}
	\label{fig:marginal_utility}
\end{figure}

@@

\begin{equation}
	\begin{aligned}
		\frac{c_d}{c_i}(P, t) = \frac{(1 - \delta(t)) R_0 \gamma S(t) I(t)}{\delta(t) n^2}
	\end{aligned}
\end{equation}

\begin{figure}
	\includegraphics[width=\linewidth]{figures/cost_fraction}
	\caption{Cost fraction $\frac{c_d}{c_i}$ of social distancing in a population $P$ of $\delta(P)$ adherence over time, based on a population of 10,000 individuals with a seed population of 1\% infected, under the assumption of an $R_0$ of 2.67 and $\gamma$ of $\frac{1}{8.5}$. The contour lines indicate the cost fraction, i.e. what fraction of the cost of social distancing $c_d$ the cost of illness $c_i$ must be in order to make not distancing a preferred strategy.}
	\label{fig:cost_fraction}
\end{figure}

@@

This risk, viewed as a function of $R_0$ and $\delta(P)$ in Figure~\ref{fig:risk_and_compliance}, indicates the somewhat counterintuitive finding that social distancing becomes more,

@@

>>>>

\hypersetup
{
  pdftitle   = {Statistical dynamics of SARS-CoV-2 as a differential game},
  pdfauthor  = {Chris von Csefalvay}
}

\title{Statistical dynamics of SARS-CoV-2 as a differential game}
\author{Chris von Csefalvay}

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/fig1-odes}
	\caption{Some quantitative solutions for the SIR model's population dynamics over values of $R_0$ between 1.5 and 6.5, and values of $\gamma$ between $1/4.6$ and $1/12.4$ over a base population of 10,000 and a seed population of 0.1\% infected initially. For each plot, $\beta$ is inferred from an $R_0$ value of 2.67, based on Liu et al. (2020),\cite{liu2020reproductive} and the $\gamma$ parameter. The susceptible population is displayed in blue, while infected/infectious cases are marked in red and recovered cases in green.}
	\label{fig:ode_solutions}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/fig3-SIR-by-delta}
	\caption{Contour diagrams of numeric solutions for susceptible, infectious and removed ($S$, $I$ and $R$) compartment sizes for a base population of 10,000 individuals with a seed population of 0.1\% infected, under the assumption of an $R_0$ of 2.67 and $\gamma$ of $\frac{1}{8.5}$. As this figure indicates, the 'critical mass' of social distancing takes place in the $\delta$ range of 0 to 0.4, and thus even modest increases in social distancing participation at low levels can make a significant difference in the number of infectious cases.}
	\label{fig:fig3-SIR-by-delta}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/marginal_utility}
	\caption{Marginal utility of social distancing in a population $P$ of $\delta(P)$ adherence over time. The marginal utility is defined as the vertical component of the gradient of infected individuals. The plot draws on a base population of 10,000 individuals with a seed population of 0.1\% infected, under the assumption of an $R_0$ of 2.67 and $\gamma$ of $\frac{1}{8.5}$.}
	\label{fig:marginal_utility}
\end{figure}

@@

\begin{equation}
	\begin{aligned}
		\frac{c_d}{c_i}(P, t) = \frac{(1 - \delta(t)) R_0 \gamma S(t) I(t)}{\delta(t) n^2}
	\end{aligned}
	\label{eq:cost_fraction}
\end{equation}

\begin{figure}
	\includegraphics[width=\linewidth]{figures/cost_fraction}
	\caption{Cost fraction $\frac{c_d}{c_i}$ of social distancing in a population $P$ of $\delta(P)$ adherence over time, based on a population of 10,000 individuals with a seed population of 0.1\% infected, under the assumption of an $R_0$ of 2.67 and $\gamma$ of $\frac{1}{8.5}$. The contour lines indicate the cost fraction, i.e. what fraction of the cost of social distancing $c_d$ the cost of illness $c_i$ must be in order to make not distancing a preferred strategy.}
	\label{fig:cost_fraction}
\end{figure}

@@

\subsection{Strategies of social distancing} % (fold)
\label{sub:strategies_of_social_distancing}

As Figure~\ref{fig:fig3-SIR-by-delta} indicates for a given $R_0$ and $\frac{\beta}{\gamma}$, social distancing can have an overwhelmingly significant effect on the number of infectious individuals in a closed population, dependent on the number of persons in the population already engaged in social distancing. This effect is most pronounced early in the epidemic (approx. 2-3 $\gamma^{-1}$ days), and the effect is most significant where less than half of the population is engaged in social distancing. Thus, unlike collective immunity in the case of vaccination, which often necessitates a fairly high level of penetration (typically estimated as $1 - R_0^{-1}$), social distancing can play a meaningful role in particular where much of the population is not engaged in such behaviours. This result can meaningfully inform a policy of encouraging individual social distancing early in an outbreak and dispelling perceptions that unless a critical volume of individuals are participating, marginal action is unnecessary.

% subsection strategies_of_social_distancing (end)

\subsection{Marginal utility of social distancing} % (fold)
\label{sub:marginal_utility_of_social_distancing}

Based on the key epidemiological dynamics data on the SARS-CoV-2 pandemic, the highest marginal effect of social distancing takes place in the same early timeframe of approx. 2-3 $\gamma^{-1}$ days. Unsurprisingly, even without integrating the time-dependent discount factor proposed by Reluga (2010), the numerical solutions indicate that the effect of social distancing is most significant where it is not yet a widely adopted strategy: for SARS-CoV-2, based on an initial population of 10,000 with a seed population of 0.1\% infected, the greatest marginal utility is encountered where less than 20\% of the population is engaged in such behaviour, and the effect is significantly less noticeable once $\delta(P)$ reaches 0.5.

Calculations of marginal utility matter because they can guide public policy in determining what fraction of the population may feasibly be exempted from social distancing. Given the need for critical services to continue even in the throes of a pandemic, marginal utility calculations based on empirical data on an outbreak itself, as well as the societal response (as expressed by $\delta(P, t)$), may contribute to more accurate public health measures while limiting the effect of such measures on the economy.

% subsection marginal_utility_of_social_distancing (end)

\subsection{Costs and strategy choices} % (fold)
\label{sub:costs_and_strategy_choices}

The costs both of social distancing and of failing to do so are notoriously difficult to quantify accurately. Indeed, many of these costs are by their very nature not amenable to quantification. At the same time, by quantifying the relative ratio of cost of distancing ($c_d$) and cost of illness ($c_i$), we can identify a strategy-associated cost relationship that, given solid information, can assist in societal decision-making as to social distancing. As Figure~\ref{fig:cost_fraction} shows, for most cases, the cost of social distancing would have to exceed the cost of illness by at least an order of magnitude to make it a preferable strategy. In addition, the computational solution of Equation~\eqref{eq:cost_fraction} shows not only that failure to socially distance may only be a preferable choice if the costs of distancing vastly exceed the costs of illness, but that this remains the case for much of the short term (<90 days).

% subsection costs_and_strategy_choices (end)

@@

\section{Discussion} % (fold)
\label{sec:discussion}

Pandemics pose a significant challenge to public health and social decision-making, and the COVID-19 pandemic is by no means an exception. Non-pharmaceutical interventions, such as social distancing, play a significant role in the public health arsenal and may play a determinative role in stemming the tide of an infectious disease that is otherwise not amenable to treatment or prophylaxis. Thus, until a vaccine or a reliable therapeutic, ideally with prophylactic properties, is found, non-pharmaceutical interventions are the mainstay of public health activity in the face of COVID-19.

This paper discussed a subject that is not devoid of controversy, both in the scientific and in the public realm. By their nature, non-pharmaceutical interventions interfere with citizens' day-to-day lives and may have complex economic, social and psychological effects. It is therefore important that strategy options are adequately explored from a quantitative perspective. It is hoped that in reinforcing the case for social distancing through an analysis of the statistical dynamics that underlie it, this paper can add to the wealth of literature in support of social distancing as an effective and cost-efficient non-pharmaceutical intervention where other tools are unavailable or inappropriate.

% section discussion (end)

\section*{Competing interests} % (fold)
\label{sec:competing_interests}

The author declares no competing interests.

% section competing_interests (end)

\section*{Supplementary data} % (fold)
\label{sec:supplementary_data}

All simulations, code and data are available on Github and under the DOI XXXXXXX.

% section supplementary_data (end)
